---
title: "WGCNA tutorial"
output:
  html_document:
    fig_width: 9
    fig_height: 7
---



WGCNA: Weighted gene co-expression network analysis
========================================================
This code has been adapted from the tutorials available at [WGCNA website](http://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/)

Installing required packages:
WGCNA requires the following packages to be installed,
currently one of them is only available through bioconductor

```{r eval=FALSE}

module load stats/R/3.2.1
install.packages(c("dynamicTreeCut", "cluster", "flashClust", "Hmisc", "reshape", "foreach", "doParallel") ) 
source("http://bioconductor.org/biocLite.R") 
biocLite("impute")
install.packages("WGCNA")

```


Loading WGCNA library, and settings to allow parallel execution


```{r messages=FALSE}

library(WGCNA)
library("flashClust")
options(stringsAsFactors = FALSE);
enableWGCNAThreads()

```


Loading the data;
WGCNA requires genes be given in the columns


```{r 'load-data'}

load("/groups/pklab/scw/scw2015/varinfo.RData");
mydata=varinfo$mat;
dim(mydata)
gene.names=names(sort(varinfo$arv,decreasing=T));
mydata.trans=t(mydata);

```


For the purpose of this exercise, we focus on a smaller set of genes:


```{r 'select-subdata'}

n=500;
datExpr=mydata.trans[,gene.names[1:n]];
SubGeneNames=gene.names[1:n];

```


Choosing a soft-threshold power: a tradeoff between scale free topology and mean connectivity


```{r 'softpower selection'}

powers = c(c(1:10), seq(from = 12, to=20, by=2));
sft=pickSoftThreshold(datExpr,dataIsExpr = TRUE,powerVector = powers,corFnc = cor,corOptions = list(use = 'p'),networkType = "signed")

# Plot the results
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;

# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit, signed R^2",type="n", main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],labels=powers,cex=cex1,col="red");

# Red line corresponds to using an R^2 cut-off
abline(h=0.80,col="red")

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")

```


Generating adjacency and TOM similarity matrices based on the selected softpower


```{r 'generate-TOM'}

softPower = 8;

#calclute the adjacency matrix
adj= adjacency(datExpr,type = "signed", power = softPower);

#turn adjacency matrix into a topological overlap matrix (TOM) to minimize the effects of noise and spurious associations
TOM=TOMsimilarityFromExpr(datExpr,networkType = "signed", TOMType = "signed", power = softPower);
colnames(TOM) =rownames(TOM) =SubGeneNames
dissTOM=1-TOM
```


Module detection


```{r 'module-detection'}

#hierarchical clustering of the genes based on the TOM dissimilarity measure
geneTree = flashClust(as.dist(dissTOM),method="average");

#plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="",cex=0.3);

# Set the minimum module size
minModuleSize = 20;

# Module identification using dynamic tree cut

dynamicMods = cutreeDynamic(dendro = geneTree,  method="tree", minClusterSize = minModuleSize);
#dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM, method="hybrid", deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = minModuleSize);

#the following command gives the module labels and the size of each module. Lable 0 is reserved for unassigned genes
table(dynamicMods)

#Plot the module assignment under the dendrogram; note: The grey color is reserved for unassigned genes
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut", dendroLabels = FALSE, hang = 0.03, addGuide = TRUE, guideHang = 0.05, main = "Gene dendrogram and module colors")


#set the diagonal of the dissimilarity to NA 
diag(dissTOM) = NA;

#Visualize the Tom plot. Raise the dissimilarity matrix to a power  to bring out the module structure
sizeGrWindow(7,7)
TOMplot(dissTOM^4, geneTree, as.character(dynamicColors))

```

Extract modules

```{r 'extract-modules'}

module_colors= setdiff(unique(dynamicColors), "grey")
for (color in module_colors){
    module=SubGeneNames[which(dynamicColors==color)]
    write.table(module, paste("module_",color, ".txt",sep=""), sep="\t", row.names=FALSE, col.names=FALSE,quote=FALSE)
    
}
```

Look at expression patterns of these genes, as they are clustered

```{r 'expression-pattern'}

module.order <- unlist(tapply(1:ncol(datExpr),as.factor(dynamicColors),I))
m<-t(t(datExpr[,module.order])/apply(datExpr[,module.order],2,max))
heatmap(t(m),zlim=c(0,1),col=gray.colors(100),Rowv=NA,Colv=NA,labRow=NA,scale="none",RowSideColors=dynamicColors[module.order])

```

We can now look at the module gene listings and try to interpret their functions; for instance using http://amigo.geneontology.org/rte

We can then quantify module similarity by eigengene correlation. Eigengenes: Module representatives

```{r 'eigengene-plot'}

#MEList = moduleEigengenes(datExpr, colors = dynamicColors)
#MEs = MEList$eigengenes
#plotEigengeneNetworks(MEs, "", marDendro = c(0,4,1,2), marHeatmap = c(3,4,1,2))

```


