---
title: "Heterogeneity analysis II"
output: html_notebook
---

Here we'll use a recent dataset on chromaffin cell differentiation to look at different visualization and pseudotime techniques.

## Loading data
We generally start with count matrices. These are obtained from alignment results and protocol-specific demultiplexing routines. Chromaffin dataset was measured using SMART-seq2 protocol, and the reads for each cell are then reported in a separate bam file. We use featureCounts to extract the read counts per gene:

`for name in ``find ./ -name \"*.bam\" -print``; do sem -j 50 featureCounts -p -a mm10.genes.gtf -o ${name%.bam}.counts $name; done;`


```{r}
counts.dir <- "../data/counts"
str(list.files(path=counts.dir))
```

We'll define a quick function to read them in in parallel:
```{r}
t.read.in.counts <- function(path,n.cores) {
  fs <- list.files(path=path,pattern=".*_unique.counts$",recursive=T)
  names(fs) <- gsub("_unique.counts","",fs)
  names(fs) <- gsub(".+?/","",names(fs))
  dat <- do.call(cbind,mclapply(fs,function(file) { df <- read.delim(paste(path,file,sep="/"),header=F,stringsAsFactors=F,skip=2); x <- df[,7]; names(x) <- df[,1]; x;},mc.cores=n.cores))
  colnames(dat) <- paste(path,colnames(dat),sep=".");
  dat
}
```

```{r}
data.path <- '../data';
counts.dir <- paste(data.path,'counts',sep='/')
dat <- t.read.in.counts(counts.dir,10)
```

```{r}

colnames(dat) <- gsub("../data/counts.","",colnames(dat))
str(dat)
rm(dat);
```

The data was the processed using scde/PAGODA:
```{r eval=FALSE}
source("pagoda.helpers.r")
library(Cairo); library(parallel)
n.cores <- 30;
min.cell.genes <- 3e3; min.cell.reads <- 1e5; min.gene.reads <- 10; min.gene.cells <- 5; min.nonfailed <- 8;
n.groups <- 10; trim <- 3;
res <- t.process.dataset(cd,nam)
res <- c(res,t.postprocess(res,name="E12_5_c",port = NULL,perplexity = 10,include.aspects = TRUE,distance.threshold = 0.9, n.clusters = 5, top.aspects = 7,return.details=T))
```


But let's read in results of the processing:
```{r}
res <- readRDS(paste(data.path,"processed.rds",sep='/'))
names(res)
```

```{r}
show.app(res$app,name='chromaffin',browse=T)
```

Let's recapitulate the t-SNE embedding plot:
```{r}
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(res$emb,col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```

Let's try different visualizations of the same dataset. 


## PCA

```{r}
# get an FPM matrix as log10(FPM+1)
#fpm <- log10(exp(scde.expression.magnitude(res$knn, res$cd))+1)
lib.size <- colSums(res$cd);
fpm <- log10(t(t(res$cd)/lib.size)+1)
```

Run PCA:
```{r}
require(pcaMethods)
base.pca <- pca(t(fpm),nPcs = 2)
```

And visualize:
```{r}
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(scores(base.pca),col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```

Let's use PAGODA library size estimation instead:
```{r}
# get an FPM matrix as log10(FPM+1)
fpm <- log10(exp(scde.expression.magnitude(res$knn, res$cd))+1)
base.pca <- pca(t(fpm),nPcs = 2)
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(scores(base.pca),col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```

We can run PCA on just a set of overdispersed genes to make things faster:
```{r}
od.genes <- names(which(res$varinfo$arv>1.2))
length(od.genes)
od.pca <- pca(t(fpm[od.genes,]),nPcs = 2)
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(scores(od.pca),col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```


## tSNE 

By default, tSNE will measure cell similarity using Euclidean distance in PCA space:
```{r}
tsne <- Rtsne(t(fpm[od.genes,]), perplexity=50, verbose = FALSE, num_threads=10) 
```

```{r}
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(tsne$Y,col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```

Another good alternative is correlation distance (on od genes, or on PCs):
```{r}
d <- as.dist(1-cor(fpm[od.genes,]))
```

```{r}
tsne2 <- Rtsne(d, is_distance = TRUE, perplexity=50, verbose = F, num_threads=10) 
```

```{r}
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(tsne2$Y,col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```

We can see what increasing or decreasing perplexity does:
```{r}
tsne3 <- Rtsne(d, is_distance = TRUE, perplexity=3, verbose = F, num_threads=10) 
```

```{r}
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(tsne3$Y,col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```

## Diffusion maps

Calculate diffusion map using overdispersed genes:
```{r}
require("destiny")
dm <- DiffusionMap(t(fpm[od.genes,]), verbose = FALSE,k=15,sigma=100)
```

visualize:
```{r}
par(mfrow=c(1,1), mar = c(0.5,0.5,0.5,0.5), mgp = c(2,0.65,0), cex = 0.9);
plot(dm@eigenvectors[,c(1,2)],col=1,bg=adjustcolor(res$cols,alpha=0.6),cex=2,pch=21,panel.first=grid(),axes=F); box();
```


## Monocle2

Monocle aims to reconstruct a tree and determine embedding that best represents that tree.
First we have to represent a dataset in a different way:
```{r}
require("monocle")
pd <- as.data.frame(cbind(res$cols ));
rownames(pd) <- colnames(fpm); colnames(pd) <- c("color")
pd = new("AnnotatedDataFrame", data = pd )

fd <- as.data.frame(ifelse(rownames(fpm) %in% od.genes,1,0));
rownames(fd) <- rownames(fpm); colnames(fd) <- c("od")
fd = new("AnnotatedDataFrame", data = fd)

fpm1 <- newCellDataSet(fpm, phenoData = pd, featureData = fd)
```

```{r}
fpm1 = setOrderingFilter(fpm1,od.genes)
mncl = reduceDimension(fpm1, max_components=2)
mncl = orderCells(mncl)
```


```{r}
plot_cell_trajectory(mncl)
plot(mncl@reducedDimS[1,],mncl@reducedDimS[2,],col=as.character(mncl@phenoData@data$col))
```

